---
title: "Discussion 1"
output: html_document
---

```{r setup, include=FALSE}
# We will usually provide a chunk like this for you. You shouldn't change this chunk until and unless you know what you are doing.

# This is setting "global chunk options"; "global" meaning applying to every chunk in the rest of the document.

#   echo = TRUE causes every chunk to appear in the knitted file.
#   warning = FALSE causes warning messages (different than errors) to NOT appear in the knitted file.
#   message = FALSE causes other messages (like those from loading packages) to NOT appear in the knitted file.

# Notice include = FALSE as a chunk option at the top of this chunk in line 6. This means that neither this chunk nor its output will appear in the knitted file.

knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Preliminaries

- This file should be in `STAT240/discussion/ds01`.

The goal of today's discussion will be to review the syllabus, think about your learning goals for STAT 240, and learn how to knit an Rmarkdown file to html.


# Problem 1

What are the dates for the first midterm, second midterm, and final exam for this course? As a reminder, dates or times for the final exam are rarely offered and require extenuating or emergency circumstances.


# Problem 2

What are your goals for taking this course and what are you hoping to learn?


# Problem 3

Use the following R chunk to compute three squared plus five.

You may find [Section 2.3.2 of the textbook ](https://bwu62.github.io/stat240-revamp/rstudio-intro.html#r-math) helpful.

```{r}
# Write your code in here!

```

Ensure that the answer is appearing in the knitted file correctly. Do not just type the answer, write the code to get the answer.


# Submission

When you have answered the above questions, please knit this file and submit **both** the R Markdown file (.Rmd) and the knitted html file (.html).

We will ask for both files for every assignment in this course.

We will grade directly from the .html file, so make sure you are satisfied with how it looks before you submit. We ask for your .Rmd as a backup in case your .html file has errors in it. 