---
title: "week2_live"
output: html_document
date: "2025-09-08"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, error = TRUE, warning = T)
```

# Week 2 lecture notes

### Mathematical operation:

-   add: +
-   subtract: -
-   multiply: *
-   divide: / (forward slash)
-   raised to: ^, **

```{r}
3+2
3-2
3*2
1:5
3/2
3^2
3**2
```

### Variables:

anything you define using the '=' or '<-' sign creates a variable in the environment

```{r}
x = 12
x <- 12
X = 12
x = 3
```

Rules for naming variables: 

- letters or alphabets 

- you can use numbers in the name, but it should always come after the letters

```{r}
1a = 2
a1 = 2
A1 = 2
```

-   you can use symbols like '\_', no'-' because this is the sign for subtract, '.'

```{r}
a1_ = 2
a1. = 2 
a2- = 2
a2 # = 2
```

Types of variables:

-   numeric: var using numbers

```{r}
class(x)
class(a1)
```

-   characters:

```{r}

x = X # here it reassigns the value of x with the value of a predefined var X
# sahifa = -1
myName = sahifa
myName = "Sahifa"
food = 'Salad'
class(myName)
```

```{r}
eq = 2+2
log_2 = log(2) 
eq = "2+2"
class(eq)
```

mathematical operations with character var:

```{r}
year = " 2025"
yr = 2025

year + 2
yr + 2
year + "2"
yr + "2"
c(year, "2")

?c
```


### Logical variable:

```{r}
t = TRUE
f = FALSE
```


Logical operations: 

- >, <, == equality, != not equal, <=, >=

```{r}
x
x < 1
x >= 1
x != 3
x == 3
```

can we do mathematical operation with T or F

```{r}
T + 2
F + 2
```

Yes, because, R considers T as 1 and F as 0.

```{r}
NA
NA + 1
log(NA)

vector = c(1, 2, 3, 4, NA, NA)
sum(vector, na.rm = T)
```

Will NA work with logical operators, espexially with == and !=

```{r}
z = NA
z == NA
is.na(z)
is.na(vector)
c()
```


# Functions

name_of_function(arg1= , arg2 = )
for help: ?name_of_function

```{r}
?sum(1,2)
sum(1, 2)
mean(c(1, 2, 3, 4))
plot(1:5, 2:6, type = "l")
```


# Vectors:

- c(): combine function

Note: c converts the vector to character even if atleast one of the variable is a character, numeric is most of it is numbers and logical variables, its logical iff you have all logical var.

```{r}
c(1, 2) + c(3, 4)

c(1, 2, 3, 3000)
x = c(1, 2, 2+2, "2+2", T, NA) +1
x
class(x)

y = c(T, 1, 2, 200, NA)
y

z = c(T, F, NA)
```


- using the ':' to create numeric vectors, creating sequences of numbers with a difference of 1 or -1 (depending on the direction specified)

```{r}
1:20
pi
pi:20
-1:-100
-1:10
10:1
exp(1):10
sqrt(2):5

"2025":2027 # if you have numbers in character format, : still works

"apple":1

T:10
```

- seq(): this will create numeric vectors 

```{r}
?seq
seq(2, 1,-1)
seq(by = -2, to = 2, from = 15)
seq(15, 2, -2)
seq(15, 2, length.out = 7)
```

## Vector Operations

#### Mathematical operations

```{r}
x = 12:20
x

y = 2:10
y

x+y
# works with all mathematical operation
```

```{r}
X = 1:5
Y = 1:9
X
Y
X+Y

Y+X

# it works with all mathematical vectors
```

```{r}
x
c(11,x)
c(x, 11)
x[10]=c(11)
x[10] = 11
x
```

Comparing two vectors to see the presence of one (second) in other (first)
```{r}
X
X >= c(2, 3)
X == c(1, 2, 3, 5)
# all of these only check for the presence of the first element in the second vector
```
For cross-checking with a list, we use the operator `%in% `

```{r}
X
X %in% c(1, 3, 5)
Y = c("Banana", "Apple", "Grapes")
Y == "Strawberry"
Y == c("Banana", "Apple", "Grapes")
Y == c("Grapes", "Strawberry")
Y %in% c("Apple", "Strawberry")
Y %in% c("Apple", "Banana")
```

# Dataframes: 

```{r}
# install.packages("tidyverse") # only one time use
library("tidyverse") # run every time you open R-Studio
```

You want to define a new dataframe, we use function `tibble()`

```{r}
data = tibble(first_5_num = 1:5, alphabets = c("a", "b", "c", "d", "e"), check_primes = first_5_num %in% c(2, 3, 5), check_vowels = alphabets %in% c("a", "e"))

data$alphabets
data
data[2]
data[1, 2]
```

### `dim()`

```{r}
dim(data)
glimpse(data)
head(data, 2)
tail(data, 2)
```

### Calling a data set

```{r}
mendota = read_csv("../../data/lake-mendota-winters-2025.csv")
```





